A standardized framework for risk-based assessment of treatment effect heterogeneity in observational healthcare databases

### 2021-05-05 submission/manuscript.rmd
typo
new version of manuscript

\vspace{10mm}
**Corresponing author**
\singlespacing 
Alexandros Rekkas, MSc

@@

More recently, “predictive” HTE analyses have been described as approaches that
provide predictions of potential outcomes in a particular patient with one
intervention versus an alternative, taking into account multiple relevant
patient characteristics [@Varadhan2013]. We focus on “risk modeling”
approaches where patients are divided into risk strata using either an existing
or an internally developed risk prediction model. Risk-stratum-specific
estimates provide an overview of the evolution of treatment effects with
increasing risk both on the relative and the absolute scale. Recently,
systematic guidance on the application of such methods has been developed
[@Kent2010; @Kent2018].

While these approaches were developed for application in randomized controlled
trials (RCTs), observational databases are also an appealing substrate.
Observational healthcare databases, such as administrative claims and electronic
health records, are already highly available for the analysis of
pharmacoepidemiologic research questions [@AdlerMilstein2017; @Dahabreh2014].
They are also often larger than many typical trials, providing excellent power
for HTE analysis, including heterogeneous populations. However, unlike RCTs,
treatment effects are subject to confounding while the unique structure of
different databases calls for database-specific analysis plans that are often
not easily transportable.

The Observational Health Data Sciences and Informatics (OHDSI) collaborative has
established an international network of data partners and researchers that aim
to bring out the value of health data through large-scale analytics by mapping
all available databases to the Observational Medical Outcomes Partnership (OMOP)
Common Data Model (CDM) [@hripcsak2015observational; @Overhage2012]. The
common data structure enables analyses at a very large scale. For example, in a
recent study, a large set of first-line treatments for hypertension was compared
with respect to 55 outcomes in a network of databases, including 4.9 million
patients from around the world [@Suchard2019].

We aimed to develop a framework for risk-based assessment of treatment effect
heterogeneity in high-dimensional observational data, which extends the existing
guidelines of the RCT setting. We implemented the framework using existing OHDSI
methods for use in the OMOP-CDM, including the patient-level prediction
framework and the population-level effect estimation framework based on new-user
cohort design [@Reps2018; @Ryan2013]. As a proof-of-concept we analyzed
heterogeneity of the effects of first-line hypertension treatment: we compared
the effect of angiotensin converting enzyme (ACE) inhibitors to beta blockers
on 9 outcomes across three different US claims databases.

@@

The proposed framework defines 5 distinct steps that enable a standardized
approach for risk-based assessment of treatment effect heterogeneity for
databases mapped to the OMOP-CDM. These are: 1) general definition of the
research aim; 2) identification of the database within which the analyses will
be performed; 3) a prediction step where internal or external prediction models
are used to assign patient-level risk predictions; 4) an estimation step where
absolute and relative treatment effects are estimated within risk strata; 5)
presentation and evaluation of the results. A overview of the procedure
can be seen in Figure \@ref(fig:figure1).

@@

The aim of this step is the estimation of treatment effects (both on the
relative and the absolute scale) within risk strata---typically 4 risk
quarters---defined using the prediction model of step 3. Any appropriate method
for the evaluation of relative and absolute treatment effects can be considered,
as long as the this is done consistently in all risk strata. Common approaches
are odds ratios or hazard ratios for the relative scale and differences in
proportions or differences in Kaplan-Meier estimates on the absolute scale,
depending on the problem at hand. We estimate propensity scores within risk
strata which we then use to match patients from different treatment cohorts
or stratify them into groups with similar propensity scores or to weigh
each patient's contribution to the estimation process [@Austin2011].

@@

In the presence of a positive treatment effect and a well-discriminating
prediction model we expect an increasing pattern of the differences in the
absolute scale, even if the relative effects remain constant on the relative
scale across risk strata. Results should be presented side-by-side on a
forest-plot-like format, so that the evolution of treatment effects across risk
strata is visible both on the relative and the absolute scale.

@@

As a proof of concept, we focus on the comparison of angiotensin converting
enzyme (ACE) inhibitors to beta blockers. ACE inhibitors are among the most
common treatment classes for hypertension, with well-established effectiveness.
Beta blockers, even though initially widely used for the treatment of
hypertension, more recent trials and meta-analyses have cast doubt on their
relative effectiveness [@wiysonge2017beta]. As a result, newer US guidelines
do not consider them for initial treatment for hypertension while in the EU
guidelines combination with other antihypertensive treatments is recommended
[@Whelton2018; @Williams2018]. However, another meta-analysis suggested that
the efficacy profile of beta blockers is similar to other major treatment
classes in younger hypertensive patients and, thus, countries like Canada still
include them as a first-line treatment candidate [@Khan2006; @Rabi2020].

@@

All cohort definitions can be found in the supplementary material.

@@

In recent years several methods for the evaluation of treatment effect
heterogeneity have been developed in the setting of RCTs [@Rekkas2020].
However, low power and restricted prior knowledge on the mechanisms of variation
in treatment effect are often inherent in RCTs, which are often adequately
powered only for the analysis of the primary outcome. Observational databases
contain a large amount of information on treatment assignment and outcomes of
interest, while also capturing key patient characteristics. Our framework
provides a standardized approach that can be used to leverage available
information from these data sources, allowing for large-scale risk-based
assessment of treatment effect heterogeneity. Multiple outcomes can be evaluated
in patient subgroups of similar baseline outcome risk. Different outcome risk
stratification schemes can also be considered. The standardized nature of the
framework enables transportability to multiple databases, provided that they are
mapped to the OMOP-CDM.

@@

Externally derived and well transportable prediction models are preferred for
analyzing treatment effect heterogeneity [@Kent2010]. In the absence of such
prediction models, simulations of RCTs have shown that internal models can be
used to provide unbiased estimates of treatment effect across the spectrum of
baseline risk [@Burke2014]. However, in observational databases treatment arms
may significantly differ in sample size. Because the prediction model will
possibly better fit to the larger treatment arm, this may introduce spurious
treatment-covariate interactions in the prediction model, leading to sub-optimal
risk stratification. As a remedy, we first match the patients in the treatment
and the comparator cohorts on the basis of propensity scores. Additionally, we
propose to assess model performance in the separate treatment arms to evaluate
its aptness for risk stratification.

>>>>

\vspace{10mm}
**Corresponding author**
\singlespacing 
Alexandros Rekkas, MSc

@@

Predictive HTE analyses have been described as approaches that
provide predictions of potential outcomes in a particular patient with one
intervention versus an alternative, taking into account multiple relevant
patient characteristics [@Varadhan2013]. We focus on “risk modeling”
approaches where patients are divided into risk strata using either an existing
or an internally developed risk prediction model. Risk-stratum-specific
estimates provide an overview of the evolution of treatment effects with
increasing risk both on the relative and the absolute scale. Recently,
systematic guidance on the application of such methods has been developed
[@Kent2010; @Kent2018].

While these approaches were developed for application in randomized controlled
trials (RCTs), observational databases are also an appealing substrate.
Observational healthcare databases, such as administrative claims and electronic
health records, are already highly available for the analysis of
pharmacoepidemiologic research questions [@AdlerMilstein2017; @Dahabreh2014].
They are also often larger than many typical trials, providing excellent power
for HTE analysis, including heterogeneous populations. However, unlike RCTs,
treatment effects are subject to confounding, while the unique structure of
different databases calls for database-specific analysis plans that often are
not easily transportable.

The Observational Health Data Sciences and Informatics (OHDSI) collaborative has
established an international network of data partners and researchers that aim
to bring out the value of health data through large-scale analytics by mapping
all available databases to the Observational Medical Outcomes Partnership (OMOP)
Common Data Model (CDM) [@hripcsak2015observational; @Overhage2012]. The
common data structure enables analyses at a very large scale. We aimed to
develop a framework for risk-based assessment of treatment effect heterogeneity
in high-dimensional observational data, which extends the existing guidelines of
the RCT setting. We implemented the framework using existing OHDSI methods for
use in the OMOP-CDM, including the patient-level prediction framework and the
population-level effect estimation framework based on new-user cohort design
[@Reps2018; @Ryan2013]. As a proof-of-concept we analyzed heterogeneity of the
effects of first-line hypertension treatment: we compared the effect of
angiotensin converting enzyme (ACE) inhibitors to beta blockers on 9 outcomes
across three different US claims databases.

@@

The proposed framework defines 5 distinct steps that enable a standardized
approach for risk-based assessment of treatment effect heterogeneity for
databases mapped to the OMOP-CDM. These are: 1) general definition of the
research aim; 2) identification of the database within which the analyses will
be performed; 3) a prediction step where internal or external prediction models
are used to assign patient-level risk predictions; 4) an estimation step where
absolute and relative treatment effects are estimated within risk strata; 5)
presentation and evaluation of the results. An overview of the procedure
can be seen in Figure \@ref(fig:figure1).

@@

The aim of this step is the estimation of treatment effects (both on the
relative and the absolute scale) within risk strata---typically 4 risk
quarters---defined using the prediction model of step 3. Any appropriate method
for the evaluation of relative and absolute treatment effects can be considered,
as long as the this is done consistently in all risk strata. Common approaches
are odds ratios or hazard ratios for relative scale estimates and differences in
observed proportions or differences in Kaplan-Meier estimates for absolute scale
estimates, depending on the problem at hand. We estimate propensity scores
within risk strata which we then use to match patients from different treatment
cohorts or stratify them into groups with similar propensity scores or to weigh
each patient's contribution to the estimation process [@Austin2011].

@@

In the presence of a positive treatment effect and a well-discriminating
prediction model we expect an increasing pattern of the differences in the
absolute scale, even if treatment effects remain constant on the relative scale
across risk strata. Results should be presented side-by-side on a
forest-plot-like format, so that the evolution of treatment effects across risk
strata is visible both on the relative and the absolute scale.

@@

As a proof of concept, we focus on the comparison of angiotensin converting
enzyme (ACE) inhibitors to beta blockers are among the most common treatment
classes for hypertension, with well-established effectiveness. Beta blockers,
even though initially widely used for the treatment of hypertension, more recent
trials and meta-analyses have cast doubt on their relative effectiveness
[@wiysonge2017beta]. As a result, newer US guidelines do not consider beta
blockers for initial treatment for hypertension while in the EU guidelines
combination with other antihypertensive treatments is recommended
[@Whelton2018; @Williams2018]. However, another meta-analysis suggested that
the efficacy profile of beta blockers is similar to other major treatment
classes in younger hypertensive patients and, thus, countries like Canada still
include them as a first-line treatment candidate [@Khan2006; @Rabi2020].

@@

All cohort definitions were identical to the ones used in the multinational
study carried out within OHDSI that provided overall treatment effect estimates
comparing all anti-hypertensive drug classes with each other [@Suchard2019].
More information can be found in the supplementary material.

@@

In recent years several methods for the evaluation of treatment effect
heterogeneity have been developed in the setting of RCTs [@Rekkas2020].
However, low power and restricted prior knowledge on the mechanisms of variation
in treatment effect are often inherent in RCTs, which are usually adequately
powered only for the analysis of the primary outcome. Observational databases
contain a large amount of information on treatment assignment and outcomes of
interest, while also capturing key patient characteristics. Our framework
provides a standardized approach that can be used to leverage available
information from these data sources, allowing for large-scale risk-based
assessment of treatment effect heterogeneity. Multiple outcomes can be evaluated
in patient subgroups of similar baseline outcome risk. Multiple outcome risk
stratification schemes can also be considered. However, this should be done with
caution, as it may hinder the interpretability of the results, in a similar
manner as typical subgroup analyses. The standardized nature of the framework
enables transportability to multiple databases, provided that they are mapped to
the OMOP-CDM.

@@

Externally derived and well transportable prediction models are preferred for
analyzing treatment effect heterogeneity [@Kent2010]. In the absence of such
prediction models, simulations of RCTs have shown that internal models can be
used to provide unbiased estimates of treatment effect across the spectrum of
baseline risk [@Burke2014]. However, in observational databases treatment arms
may significantly differ in sample size. Because the prediction model will
possibly better fit to the larger treatment arm, spurious treatment-covariate
interactions may be introduced in the prediction model, leading to sub-optimal
risk stratification. As a remedy, we first match the patients in the treatment
and the comparator cohorts on the basis of propensity scores. Additionally, we
propose to assess model performance in the separate treatment arms to evaluate
its aptness for risk stratification.
